---
title: "basicFunctions"
output: github_document
date: "2026-10-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Where we are
- load data into R (read.table())
- work with vectors, factors and data frames
- access elements with [ ] and $
- make first plots with the standard library and with ggplot2

## Today we learn
- write own functions
- automate repetitive work with loops, sapply and replicate
- make decisions with if/else
- work with matrices and matrix multiplication
- explore an interesting data set: how much do mammals sleep
- more essentials of the standard graphics library
- learn more ggplots


# Functions and apply

## simple functions; single parameter
Lets start with the most simple one
```{r}
square<-function(a)
{
  s=a*a
return (s)
}

# lets use it
test<-square(5)
print(test)
```
It also works with irregular parameters
```{r}
test<-square(TRUE)
print(test)
```

## simple functions; multiple parameters
```{r}
rectangle<-function(a,b)
{
  a=a*b
return (a)
}

# lets use it
ra<-rectangle(5,4)
print(ra)
```

## simple functions; multiple rectangles

Now we want to compute the area of multiple rectangles. How to best do this?
```{r}
a<- c(10,50,80,90)
b<- c(40,20,30,40)
```


## Classically with a for loop

```{r}
# preallocate the vector with the results:
area <- numeric(length(a))
for (i in 1:length(a)) {
  area[i] <- rectangle(a[i], b[i])
}
print(area)
```

## vectorization
But actually R is vectorized, hence we can do it much simpler!
```{r}
area <-  rectangle(a, b)
print(area)
```

## can all functions be vectorized?
What about this one:
```{r}
# find the longer side of a rectangle a, b
longer_side <- function(a, b) {
  max(a, b)
}
```

Lets try with two values
```{r}
print(longer_side(10,30))
print(longer_side(35,1))
```


now with our vector
```{r}
a<- c(10,50,80,90)
b<- c(40,20,30,40)
print(longer_side(a,b))

```

How to deal with this problem?

- Solution 1: loop - see above

- Solution 2: mapply (multi apply)

## mapply
```{r}
long<-mapply(longer_side,a,b)
print(long)
```

## sapply and if/else
I want to know if i sleep enough.
So i make a function giving me the answer

```{r}
sleep_enough <- function(hours) {
  var<-""
  if (hours >= 14) {
    var<-"more than enough"
  } else if (hours >= 8) {
    var<-"enough"
  } else {
    var<-"short"
  }
  return(var)
}
```

Do I sleep enough
```{r}
# Lets see, do i sleep enough?
print(sleep_enough(7))
```
can i use this function with multiple people?
```{r}
mp<-c(1,7,8,14,12,1)
#print(sleep_enough(mp))
# no i get an error! 
```
One solution is sapply
```{r}
mp<-c(1,7,8,14,12,1)
res<-sapply(mp,sleep_enough)
print(res)
```
**To summarize: apply applies a function to single (sapply) or multiple (mapply) values**

## replicate
Similar to sapply, allows you to execute a function a given number of times; it is only useful for random processes as otherwise you always get the same results

```{r}
rep<-replicate(5,square(20))
print(rep)
rep<-replicate(5,rnorm(1))
print(rep)

# this was just for demonstration; because actually...
rep<-rnorm(5)
print(rep)
```

# Matrices
## generate a simple 2x2 matrix

```{r}
# we need to start with a vector
vec<-c(1, 2, 3, 4)
print(vec)
# and distort it into a matrix
m<-matrix(vec,nrow=2)
print(m)
```
We can also generate the matrix by row
```{r}
# by row
m<-matrix(vec,nrow=2,byrow=TRUE)
print(m)
```
## subsets of matrices
```{r}
m<-matrix(c(1, 2, 3, 4),nrow=2)
print(m)
print(m[1,2])
print(m[2,1])
```
Or we can print entire rows:
```{r}
print(m[1,])
print(m[2,])
```
Or print entire columns:
```{r}
print(m[,1])
print(m[,2])
```
## What are matrices? 
### to illustrate the transforming powers of matrices lets make a letter F based on coordinates
```{r}

fletter <- matrix(c(0, 0,   1, 0,   1, 2,   2.5, 2,   2.5, 3,
                  1, 3,   1, 4,   3, 4,   3, 5,     0, 5), nrow = 2)
print(fletter)
```

How is this an F? lets visualize with our favorite ggplot2
```{r}
library(ggplot2)
# geom polygon, provide all points of a polygon, and the points will be connected like in the children game, where you connected the dots
df<-data.frame(x = fletter[1, ], y = fletter[2, ])
g<-ggplot(df, aes(x = x, y = y)) +
  geom_polygon(fill = "steelblue", alpha = 0.6) + coord_equal()
plot(g)
```
### so what is a matrix doing to our F-letter?
```{r}
scalematrix <- matrix(c(2, 0, 0, 0.5), nrow = 2, byrow = TRUE)
print(scalematrix)
# matrix multiplication %*%
scaledletter <- scalematrix %*% fletter
print(scaledletter)

```
visualize scaled matrix
```{r}
df <- data.frame(x = c(fletter[1, ], scaledletter[1, ]),
                 y = c(fletter[2, ], scaledletter[2, ]),
                 version = rep(c("original", "transformed"), each = ncol(fletter)))
print(df)
```


```{r}
g<-ggplot(df, aes(x = x, y = y, fill = version)) +
  geom_polygon(alpha = 0.5) +
  coord_equal()
plot(g)
```
### now lets turn this into a function


```{r}
transform<-function(letter,matrix)
{
  # first transform
  trans<-matrix %*% letter
  # second store in dataframe
  df <- data.frame(x = c(letter[1, ], trans[1, ]),
                 y = c(letter[2, ], trans[2, ]),
                 version = rep(c("original", "transformed"), each = ncol(fletter)))
  # third visualize with ggplot2
  g<-ggplot(df, aes(x = x, y = y, fill = version)) +
    geom_polygon(alpha = 0.5) +
    coord_equal()
  plot(g)
}
```

### lets transform our letter with multiple matrices

#### identity
```{r}
id<-matrix(c(1, 0, 0, 1),nrow=2)
print(id)
```

```{r}
transform(fletter,id)
# no change => thats why this is called identity matrix
```



#### transposing
```{r}
tp<-matrix(c(0, 1, 1, 0),nrow=2)
print(tp)
```

```{r}
transform(fletter,tp)
# transposing = exchange x and y-letters
```
#### x-scaling
```{r}
xs<-matrix(c(0.5, 0, 0, 1),nrow=2)
print(xs)
```

```{r}
transform(fletter,xs)
# x-scaling
```
### What to do when i want to i) transpose and x-scale? Matrix multiplication!

#### attempt one - the naive approach


```{r}
dt1<-tp*xs
print(dt1)
```
```{r}
transform(fletter,dt1)
# desaster, my polygon has now shrunk to a point at coordinates 0
```
#### attempt two - proper matrix multiplication


```{r}
dt2<-xs %*% tp
print(dt2)
```
```{r}
transform(fletter,dt2)
# voila, we have combined two linear transformation into a single matrix!
# transposing and x-scaling
```
# Explore interesting data set - how much do mammals sleep

**new Rmarkdown sleepStorry**

