---
title: "sleep-story"
output: github_document
date: "2026-10-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# the Dataset


```{r}
library(ggplot2)
print(msleep) # comes with ggplot2
```
```{r}
# it's a tidyverse tibble
# tidyverse and tibbles are great; but we have no time to cover tidyverse (unfortunatelly)
# so lets convert to our favorite data fram
ms <- as.data.frame(msleep)
# lets look at the data with str()
str(ms)
```
## what are the columns

- name: common name of the species
- genus, order: taxonomy
- vore: diet; carni(vore), herbi(vore), insecti(vore), omni(vore)
- conservation: IUCN conservation status (lc = least concern, vu = vulnerable, en = endangered, …)
- sleep_total: total sleep (hours per day)
- sleep_rem: REM sleep (hours per day)
- sleep_cycle: length of one sleep cycle (hours)
- awake: hours awake per day (= 24 - sleep_total)
- brainwt: brain weight (kg)
- bodywt: body weight (kg)


# Lets explore various questions
## Who sleeps the most?

```{r}
print(max(ms$sleep_total))
```

Fantastic - but who actually is 19.9 hours?
Any guesses? 

```{r}
which.max(ms$sleep_total)
# so the index of the winner is...
```
```{r}
ms[which.max(ms$sleep_total),]
# the actual winner is
```
## Whats the sleep distribution

using the standard library hist (histogram)
```{r}
hist(ms$sleep_total)
```
## Is there a relationship between sleep and body weight?

```{r}
plot(log10(ms$bodywt), ms$sleep_total, pch = 19,
     xlab = "log10 body weight (kg)", ylab = "sleep length in hours")
```
```{r}
ms$lbw<-log10(ms$bodywt)
plot(ms$lbw, ms$sleep_total, pch = 19,
     xlab = "log10 body weight (kg)", ylab = "sleep length in hours")
fit <- lm(sleep_total ~ lbw , data = ms)
abline(fit, col = "red", lwd = 2)
```

So thats interesting: the more heave a mammal the less it sleeps

## Lets use the standard library to look at several aspects of the data
```{r}
par(mfrow = c(2, 2)) # 2 rows, 2 columns
hist(ms$sleep_total, main = "Total sleep", xlab = "hours")
hist(ms$sleep_rem, main = "REM sleep", xlab = "hours")
hist(log10(ms$bodywt), main = "Body weight", xlab = "log10 kg")
hist(log10(ms$brainwt), main = "Brain weight", xlab = "log10 kg")
```
# Now lets look at the data with ggplot

## Refresher ggplot (grammar of graphics):

A ggplot is built from layers that are combined with +:

- data: a data frame
- aesthetics aes(): map variables to visual properties (x, y, color, fill, size, shape, group, label, …)
- geoms: what to draw (geom_point(), geom_line(), geom_boxplot(), …)
- optionally scales, facets and themes


### Brain weight vs bodyweight

```{r}
ggplot(ms, aes(x = bodywt, y = brainwt)) +
  geom_point() 
```
Nice - but we do not really see a trend. Can we do better, e.g. by making the axis log10.

```{r}
ggplot(ms, aes(x = bodywt, y = brainwt)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10()
```
There is a clear trend - the more body weight the more brain weight!
### I have a hypothesis, could this depend on the diet?
```{r}
ggplot(ms, aes(x = bodywt, y = brainwt,color=vore)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10()
```

### Lets add a regression line

```{r}
ggplot(ms, aes(x = bodywt, y = brainwt,color=vore)) +
  geom_point() +
   geom_smooth(method = "lm", formula = y ~ x,color="black") +
  scale_x_log10() +
  scale_y_log10()+
    labs(x = "Body weight (kg)", y = "Brain weight (kg)", color = "Diet")
```
### lets highlight some particular interesting species

```{r}
highlight <- subset(ms, name %in% c("Human", "Chimpanzee", "African elephant",
                                    "House mouse", "Cow"))
ggplot(ms, aes(x = bodywt, y = brainwt)) +
  geom_point(color = "grey60") +
  geom_point(data = highlight, color = "red", size = 2) +
  geom_text(data = highlight, aes(label = name), vjust = -1, size = 3) +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "Body weight (kg)", y = "Brain weight (kg)")
```

### Hypothesis: do carnivors need more sleep

```{r}
ggplot(ms, aes(x = vore, y = sleep_total)) +
  geom_boxplot()
```
I do not like that the ones without info are a separate group lets remove them

```{r}
msv <- subset(ms, !is.na(vore)) # ! means NOT
nrow(msv)
ggplot(msv, aes(x = vore, y = sleep_total)) +
  geom_boxplot()
``` 
It is sometimes argued that boxplots are not great, as they may hide outliers, so it could be helpful to actually show the data points as well
```{r}
ggplot(msv, aes(x = vore, y = sleep_total)) +
  geom_boxplot(outlier.shape = NA) + # hide outliers, the jitter shows all points anyway
  geom_jitter(width = 0.2, alpha = 0.5) +
  labs(x = "Diet", y = "Total sleep (hours per day)")
```

## Faceting - how does sleep and body_weight correlate for different 'vora'

Either showing colors
```{r}
ggplot(msv, aes(x = bodywt, y = sleep_total,color=vore)) +
  geom_point() +
  scale_x_log10() +
  labs(x = "Body weight (kg)", y = "Total sleep (hours per day)")
```
Or show them separately using faceting
```{r}
ggplot(msv, aes(x = bodywt, y = sleep_total)) +
  geom_point() +
  scale_x_log10() +
  facet_wrap(~ vore) +
  labs(x = "Body weight (kg)", y = "Total sleep (hours per day)")
```

# Assignment until next week
* Do the same with a dataset of your choice. 
* Tell me a scientific story in RMarkdown; generate a pdf of the story including visualizations with ggplot2 and sent them to biomedpython@gmail.com by end of next week.
* Be creative and ask questions; tell an interesting story with the analysis (not using excessive words)
* You could do it with the lungdeaths; but feel free to use any data set of interest (voting behaviour, genetic data, economic development, healthcare, etc)